Interpretable Neural Networks using EAGGA

### 2025-03-14 paper/main.tex
interpretability taxonomy

% Taxonomy
\citet[chap. 2]{survey_NN_interpretability} extend this distinction to a three-dimensional taxonomy, allowing for better categorisation of neural networks (NNs),
a model class that in its fully-connected feedforwad form is inherently non-interpretable.
- \textbf{Passive vs Active Approaches}
  * passive: post-hoc
  * active: changing architecture or training process to make model interpretable
- \textbf{Type of Explanations}
  * examples: providing examples of what leads to desired output
  * attribution: attributing effect on output for a specific feature
  * hidden semantics: examine what type of inputs specific neurons / layers pick up on
  * rules: logical rules, e.g. if-then, trees
- \textbf{Local vs Global Interpretability}
  * local: explanation based on individual samples
  * semi-local: explanation based on set of samples, e.g. grouped together by some criterion
  * global: explaining network as a whole

(Evaluation)
- objective evaluation difficult, as no clear definition of interpretability
- \citet[3]{DoshiVelez2017TowardsAR} propose a taxonomy to categorise possible evaluation methods
  * \textbf{application-grounded evaluation}
    - most rigorous method, also most expensive + time consuming
    - idea: evaluate ``interpretability model'' directly w.r.t. the task -> human experts evaluate the outcome
    - e.g.: model performing medical diagnosis on patients would be evaluated based on doctors doing the same
  * \textbf{human-grounded metrics}
    - similar to application-grounded, but tries to simplify task (while preserving its essence) so that laypersons can do it
    - cheaper due to larger, less qualified subject pool
    - especially suitable if only general concepts of the tasks need to be validated
    - examples for evaluation set-up
      * binary forced choice: human evaluator chooses which of two explanations he finds better
      * forward simulation: evaluator must correctly simulate model output when presented model input + explanation % TODO: perhaps binary forced choice is already enough to include, this (and the third one is a bit awkward to explain why this is good for interpretability imo)
  * \textbf{functionally-grounded evaluation}
    - assess explanation quality via some formally defined proxy for interpretability
    - no human time nor cost required beyond initial formulation of proxy
    - e.g. if we already have an interpretable model class (e.g. identified via human-grounded evaluation), can then rank different models of that class based on proxy
    - main challenge: what proxies

>>>>

% Taxonomy
\citet[chap. 2]{survey_NN_interpretability} extend this distinction to a three-dimensional taxonomy, allowing for better categorisation of neural networks (NNs),
a model class that in its fully-connected feedforwad form is inherently non-interpretable.\\ % TODO: letztes halbsatz vllt lieber in intro unterbringen, hier ist der awkward und unnötig ausschmückend
\textbf{Passive vs Active Approaches}, where \textit{passive} are all post-hoc methods and \textit{active} methods actively change either the architecture or
training process to increase model interpretability.\\
\textbf{Type of Explanations}, distinguishing between \textit{example} methods, providing concrete examples of what leads to a desired output, \textit{attribution}
methods that attribute the effect on the output for a specific feature, \textit{hidden semantic} methods, which explain the types of inputs particular neurons or layers
pick up on, and logical \textit{rules}, such as if-then clauses or tree-induced rules.\\
\textbf{Local vs Global Interpretability}, ranging from \textit{local} methods providing explanations based on individual samples, \textit{semi-local} methods,
explaining model behaviour for sets of samples grouped by some criterion, to \textit{global} methods, which explain the network as a whole.

% Evaluation
- objective evaluation difficult, as no clear definition of interpretability
- \citet[3]{DoshiVelez2017TowardsAR} propose a taxonomy to categorise possible evaluation methods
  * \textbf{application-grounded evaluation}
    - most rigorous method, also most expensive + time consuming
    - idea: evaluate ``interpretability model'' directly w.r.t. the task -> human experts evaluate the outcome
    - e.g.: model performing medical diagnosis on patients would be evaluated based on doctors doing the same
  * \textbf{human-grounded metrics}
    - similar to application-grounded, but tries to simplify task (while preserving its essence) so that laypersons can do it
    - cheaper due to larger, less qualified subject pool
    - especially suitable if only general concepts of the tasks need to be validated
    - examples for evaluation set-up
      * binary forced choice: human evaluator chooses which of two explanations he finds better
      * forward simulation: evaluator must correctly simulate model output when presented model input + explanation % TODO: perhaps binary forced choice is already enough to include, this (and the third one is a bit awkward to explain why this is good for interpretability imo)
  * \textbf{functionally-grounded evaluation}
    - assess explanation quality via some formally defined proxy for interpretability
    - no human time nor cost required beyond initial formulation of proxy
    - e.g. if we already have an interpretable model class (e.g. identified via human-grounded evaluation), can then rank different models of that class based on proxy
    - main challenge: what proxies
